The role of advection versus diffusion in intraventricular transport

### 2025-03-12 paper/main.tex
changes to Introduction

% I1: What is the problem - from big picture to focus
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{CITATTION}. Disturbance of the
CSF circulation can lead to brain disorders, such as hydrocephalus
where accumulation of CSF often requires emergency surgical
treatment~\cite{CITATTION}.

% MER: Add this to the paragraph above.
%
% This is particularly important since ventricular enlargement, known
% as hydrocephalus, is observed in animal models and humans with
% motile cilia deficiency, indicating that the cilia play a crucial
% role for healthy brain functioning.

% I2: What do we know about the problem?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia lining the ventricular walls, CSF secretion by the choroid
plexus, and pulsatile pressure gradients generated by the cardiac
cycle, respiration, neural activity, and bodily
movement~\cite{CITATION}. Motile cilia are microscopic hair-like
structures that beat in a metachronal wave pattern at frequencies of
10--30 Hz~\cite{CITATION}. Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes to flow near the
ventricular surface~\cite{CITATION}. On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at scale in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
  causemann2025insilico, fultz2019coupled}.

%How the cilia contribute to overall CSF flow and solute transport
%remains poorly understood however.

%These questions have been challenging to tackle due to technical
%limitations.

% I3: What do we not know about the problem and what do we want to
% know?

%As a result, cilia dynamics are technically challenging to monitor
%in-vivo.
%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%
\mer{MER: Let's look at this next paragraph together. It should clearly lay out what we do not know about the problem (and what we want to find out). Be precise. Refer to specific studies and what specifically we (do not) know. What we know belongs nicely in the paragraph above.}
For instance, studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction will alter CSF composition and lead to brain
disorders. The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

@@

\begin{figure}%[hbt!]
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 25\% into the cardiac cycle. %t=0.112613 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{i.} Streamlines simulated with the baseline flow model at
    a time 75\% into the cardiac cycle. %t=0.315315 s
    \textbf{j.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of panels \textbf{f, h--j} are indicated.
    }
    \label{fig:fig1}
\end{figure}

@@

@@

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
by initially using fTetWild~\cite{Hu2020FastWild} and subsequently
local refinement functionality in~\cite{dolfinx2023preprint}. The mesh
consists of 141 512 tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (3.1 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.

@@

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments. 
To impose a smooth initial condition, we initially solve the diffusion equation
\begin{equation*}
    \pdifft{c} = \Tilde{D}\nabla^2 c
\end{equation*}
with an increased diffusion coefficient $\Tilde{D}=D\times 10^{5}$ for two timesteps.
The solutions are normalized by their respective maximum values, and subsequently
multiplied with the value of \Cref{eq:photoconversion_curve} at the times $t=\Delta t$
and $t=2\Delta t$, respectively.
The resulting concentration fields are used as initial conditions for $c(\xx, 0)$ and
$c(\xx, \Delta t)$ for all $\xx\in\Omega$, where $\Delta t$
is the timestep. The two initial conditions are required because of the
two-step temporal discretization scheme employed (see~\Cref{subsec:num_approx_adv_diff} for more details.)

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricular geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle to yield $\overline{U} = 2.4$ \textmu m/s.
To assess the assumption that inertial forces are negligible compared to
viscous forces, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r = 110$ \textmu m and $U_r = 27.7$ \textmu m/s
as length and velocity scales characteristic of the flow.

>>>>

% I1: What is the problem - from big picture to our precise topic - why is it interesting?
Brain development and function depend on numerous regulatory
processes, including the flow of cerebrospinal fluid
(CSF)~\cite{CITATTION}. CSF is a clear liquid mostly produced by the choroid
plexus that flows through the ventricular system and subarachnoid
space~\cite{Damkier2013}. By transporting growth factors, nutrients, neurotransmitters
and metabolites as well as removing waste products, CSF regulates
brain development and homeostasis~\cite{CITATTION}. Disruption of the
CSF circulation is detrimental to brain health and disturbs cognitive and motor functions~\cite{Johanson2008}.
Pathological CSF flow may also lead to
disorders such as hydrocephalus, where accumulation of
CSF often requires emergency surgical
treatment, but for which there is 
little agreement on optimal treatments~\cite{flannery2014pediatric}. 
Of particular importance is the fact that
hydrocephalus is observed in animal models and humans with deficient motile
cilia~\cite{Afzelius2004Cilia, Olstad2019CiliaryDevelopment, Mitchison2017Motile},
indicating that the cilia play a crucial role for healthy brain functioning.

% I2: What do we know about the problem and the topic?
The flow of CSF in the ventricular system is driven by multiple
physiological processes, including via the beating motion of motile
cilia lining the ventricular walls, CSF secretion by the choroid
plexus, and pulsatile pressure gradients generated by the cardiac
cycle, respiration, neural activity, and bodily
movement~\cite{Vinje2019RespiratoryMeasurements,
Kurtcuoglu2005ComputationalSystem, Olstad2019CiliaryDevelopment}.
Motile cilia are evolutionarily conserved, microscopic hair-like
structures that beat in a metachronal wave pattern at frequencies of
10--30 Hz to mediate fluid flow~\cite{mitchell2007evolution, Ringers2020Role}.
Given their microscopic size, ranging from
6 to 15 microns, the cilia clearly contributes primarily to flow near the
ventricular surface in the human brain~\cite{Siyahhan2014FlowVentricles}.
On the other hand, cardiac,
respiratory and delta waves at lower frequencies induce pulsatile flow
at the larger scales in the brain ventricular system, subarachnoid space, and
subarachnoid perivascular
spaces~\cite{Vinje2019RespiratoryMeasurements, eide2024functional,
Causemann2025, fultz2019coupled}.

% I3: What do we not know about the problem and what do we want to
% know? What are the precise research questions we want to address?
% Refer to specific studies and what specifically we do not know.

%
%Until now, most studies have analyzed one physical mechanism at a
%time, especially in mammals.
%

Still an underexplored topic, how the motile cilia contribute to overall CSF flow
and solute transport remains poorly understood. Experimental endeavors are
challenging to pursue due to technical limitations: the microscopic size
and high beating frequencies of the cilia make them difficult to monitor 
in vivo in humans. Studies on animal models such as mice and
zebrafish have provided unequivocal evidence that cilia are crucial
for CSF flow in these animals~\cite{Olstad2019CiliaryDevelopment, Faubel2016Cilia}.
It is clear that cilia affects advective intraventricular transport 
to some extent. However a quantification of this effect, as well as of to which
extent diffusion plays a role in transporting solutes, remains
yet to be performed.

Studies using brain explants have convincingly shown
that cilia generate complex flow patterns at the surface of the brain
ventricles; however, they failed to identify the role of pressure
gradients associated with respiration and cardiac pulsations in CSF
flow and the accompanying transport. Similarly, while MRI technology
can measure CSF movement in humans, it cannot detect the microscopic
movements of cilia to understand their impact in the large human
brain. Consequently, we lack a comprehensive understanding of how
cilia contribute to CSF motion and solute transport, and whether cilia
dysfunction will alter CSF composition and lead to brain
disorders. The technical limitations of studying cilia-mediated flow and transport opens the
possibility of mathematical modeling and computational investigations. Numerous previous studies
have focused on modeling the interaction between single or several cilia and surrounding
fluids~\cite{Guo2020SimulatingGeometries, Ruvalcaba2021NumericalTree, Smith2009MathematicalFluids,
Cui2019NumericalMethod, Cui2022AFlow}.
Others considered applications such as ciliary transport of mucous and particles in
airways~\cite{Fulford1986Muco-ciliaryLung, Ramirez-SanJuan2020Multi-scaleArrays},
propulsion and locomotion of ciliated organisms~\cite{BLAKE1974MechanicsMotion,
Jahn1972LocomotionProtozoa, Brennen1977FluidFlagella},
oxygen regulation in coral reefs~\cite{Pacherres2022CiliaryProduction},
and cilia-mediated flow in the brain~\cite{Siyahhan2014FlowVentricles,
Yoshida2022EffectVentricles, Salman2022ComputationalEmbryo, Thouvenin2020OriginCanal}.
These simulation studies have provided valuable insights on how cilia mediate complex
biological flows, and how cilia may contribute to tissue physiology.

@@

@@

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{graphics/figure1.png}
    \caption{
    \textbf{a.} Schematic illustration of the two cerebrospinal fluid (CSF) flow components modeled,
    motile cilia and cardiac pulsations, and their contributions to the flow patterns.
    \textbf{b.} Particle tracking data visualizing flow directionality in the middle ventricle.
    \textbf{c.} Flow velocity magnitudes in the middle ventricle from particle tracking data.
    \textbf{d.} Computational mesh with marked facets indicating regions of motile cilia (magenta) and
    anterior/posterior facets (blue).
    \textbf{e.} Streamlines of the (steady-state) CSF flow field simulated with the
    cilia-only/no-cardiac flow model.
    \textbf{f.} Streamlines of the CSF flow field simulated with the cardiac-only/no-cilia flow
    model at a time 25\% into the cardiac cycle. %t=0.112613 s, period=0.45045 s
    \textbf{g.} Streamlines simulated with the baseline flow model at
    a time 25\% into the cardiac cycle. %t=0.112613 s
    \textbf{h.} Streamlines simulated with the baseline flow model at
    a time 50\% into the cardiac cycle. %t=0.225225 s
    \textbf{i.} Streamlines simulated with the baseline flow model at
    a time 75\% into the cardiac cycle. %t=0.315315 s
    \textbf{j.} Illustration of the cardiac-related normal pressure forcing imposed on the
    anterior/posterior boundaries. Time instants of panels \textbf{f, h--j} are indicated.
    }
    \label{fig:fig1}
\end{figure}

@@

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
by initially using fTetWild~\cite{Hu2020FastWild} and subsequently
local refinement functionality in DOLFINx~\cite{dolfinx2023preprint}. The mesh
consists of 141 512 tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (3.1 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.

@@

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments. Because steep gradients
in the concentration field may arise from imposing $c(\xx, t)$ in $\Omega_{\mathrm{pc}}$,
the mesh is locally refined around this region.

In order to impose a smooth initial condition, we initially solve the diffusion equation
\begin{equation*}
    \pdifft{c} = \Tilde{D}\nabla^2 c
\end{equation*}
with an increased diffusion coefficient $\Tilde{D}=D\times 10^{5}$ for two timesteps.
The solutions are normalized by their respective maximum values, and subsequently
multiplied with the value of \Cref{eq:photoconversion_curve} at the times $t=\Delta t$
and $t=2\Delta t$, respectively.
The resulting concentration fields are used as initial conditions for $c(\xx, 0)$ and
$c(\xx, \Delta t)$ for all $\xx\in\Omega$, where $\Delta t$
is the timestep. The two initial conditions are required because of the
two-step temporal discretization scheme employed (see~\Cref{subsec:num_approx_adv_diff} for more details.)

@@

To evaluate the relative importance of advection and
diffusion as transport mechanisms, we consider the Péclet number
\begin{equation*}
    \mathrm{Pe} = \frac{\mathrm{diffusion \ timescale}}{\mathrm{advection \ timescale}}
    = \frac{t_d}{t_a} = \frac{L_p^2/D}{L_p/U_p} = \frac{L_p U_p}{D}.
\end{equation*}
Thus, if $\mathrm{Pe} > 1$, transport is dominated by advection,
and if $\mathrm{Pe} < 1$ transport is dominated by diffusion.
Here, $L_p$ and $U_p$ are characteristic length and velocity scales,
respectively, and $D$ is the solute diffusion coefficient.
In calculating a global Péclet number, we use $L_p=600$ \textmu m, the approximate length of the
ventricular geometry along the rostrocaudal ($x$-) axis.
We use a mean velocity $\overline{U}$ as the characteristic velocity $U_p$.
The mean velocity $\overline{U}$ is calculated by first averaging the velocity in space:
\begin{equation*}
    \overline{u}^2 = \frac{1}{|\Omega|}\int_{\Omega}\uu\cdot\uu\dx,
\end{equation*}
and then averaging $\overline{u}$ in time over one cardiac cycle to yield $\overline{U} = 2.4$ \textmu m/s.
To assess the assumption that inertial forces are negligible compared to
viscous forces, we also compute the Reynolds number
\begin{equation*}
    \mathrm{Re} = \frac{\mathrm{inertial \ forces}}{\mathrm{viscous \ forces}} = \frac{\rho U_r L_r}{\mu},
\end{equation*}
with $L_r = 110$ \textmu m (the height of the middle ventricle) and
$U_r = 27.7$ \textmu m/s (the maximum velocity magnitude)
as length and velocity scales characteristic of the flow.
